---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# actmobilityr

The goal of `actmobilityr` is to provide functions to estimate the economic value of the health effects associated to a shift from passive to active mobility in Italian municipalities. The economic valuation considers effects on
mortality only but future versions of this package will allow an estimation of effects on morbidity too.
A shift from passive to active mobility implies an increase in physical activity in new adopters, which can, in turn, trigger positive effects on health. Since most commuting is carried out in urban areas, however, such a shift can have also negative effects on health because of higher air pollutant intake and increased risk injuries. These are taken into account in the estimation.
The package allows to estimate unit benefit factors associated to a mobility shift for any of the Italian municipalities using - when available - municipality-specific air pollutant concentration data, and province-specific mortality rate and life expectancy data. The algorithm used to carry out the estimation takes into account pre-existing physical activity levels among the population (data refer to Italy as a whole but are age-specific) since health benefits due to increased physical activity are stronger for inactive people and are capped at a certain threshold of weekly physical activity, above which no additional benefits are yielded.
Active mobility transportation modes considered are walking, bikes and e-bikes.
More information on the methodology used can be found in a [paper by Mela and Girardi](https://doi.org/10.1016/j.jth.2022.101487).

## Installation

You can install the development version of actmobilityr from [GitHub](https://github.com/) with:

```{r installation, echo=TRUE, eval=FALSE}

install.packages("devtools")
devtools::install_github("giuliomela/actmobilityr")

```

```{r}

library(actmobilityr)

```


## Examples

The package currently has two functions available for the user. One, `scenario_builder`, is designed to define a mobility change scenario of which the health impacts need to be estimated. The other, `active_mob_impact`, quantifies the health benefits in economic terms.

The function `scenario_builder` can be used to create a modal shift scenario for any of the Italian municipalities. The function takes several parameters and returns a data frame containing the number of individuals shifting from a given passive mobility mode to an active one. The user can select a city of interest amongst all Italian municipalities. The full documentation can be displayed typing `?scenario_builder` in the console. One of trickiest parameters (and one of few with no default values) the user must specify is `mode_change_distance` which is a vector of length `4`. Each element of `mode_change_distance` represents the share of total commuters shifting from passive to active mobility for commuting travels shorter than 3 km (one way), between 3 and 10 km, between  10 and 15 km and the longer than 15 km respectively (commuting distances are grouped in four classes and not on a linear scale because the starting dataset - the Italian Commuting Matrix - provides the data in this forma only). The user can adjust these shares for different age groups with the `mode_change_age` parameter (the default option is that shares are the same across age groups). The output of `scenario_builder` is a data frame that is also the main input for the `active_mob_impact` function.

```{r scenario-builder-example, echo=TRUE}
library(actmobilityr)

### defining a mobility-change scenario for the city of Florence

# share of total commuters shifting to active mobility (less than 3 km, 3-10 km, 10-15 km, more than 15 km)
mode_change_shares <- c(0.2, 0.15, 0.1, 0.05) # for example, 0.2 means that 20% of total commuters travelling with a give passive mode of transportation shift to an active mode.

scenario <- scenario_builder(city = "Palermo", # city of interest
                 mode_from = "private_car_driver", # passive transportation mode
                 mode_to = "bike", # active transportation mode
                 mode_change_distance = mode_change_shares, # share of new adopters on total commuters
                 mode_change_age = rep(0, 9),
                 comm_days = 4, # weekly commuting days
                 max_km = 10) # maximum distance above which commuters don't change their transportation mode

head(scenario, 10)
```

Function `active_mob_impact` performs the economic estimation of the benefits associated with the modal shift. The main input of this function is the scenario built with `scenario_builder`, all the other parameters have defaults but can indeed be changed by the user. See `active_mob_impact` for more details.

```{r active-mob-impact-example, echo=TRUE}

active_mob_impact(scenario)


```

Function `active_mob_impact` returns a tibble with the outcome of the economic evaluation: avoided deaths and their economic value estimated  with both the VOLY and the VSL. Values per km travelled are provided if the `detail` parameters is set to `FALSE`
